---
title: "Reproducible Research: Peer Assessment 1: Scot J Matkovich"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r load_preprocess, echo = TRUE}
fileURL = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
destfile = "activity.zip"
if(!file.exists("activity.csv")) {
  download.file(fileURL, destfile)
  unzip(destfile)
}
activity <- read.csv(file="activity.csv")
activity_valid <- na.omit(activity)
```

## What is the mean and median total number of steps taken per day?

```{r mean_median_steps, echo = TRUE}
library(dplyr)
steps <- activity_valid %>% group_by(date) %>% summarize(mean_steps = mean(steps), median_steps = median(steps))
hist(steps$mean_steps, main="mean steps per day", xlab="mean steps")
print(steps, n=nrow(steps))
```

## What is the average daily activity pattern?

```{r daily_activity_pattern, echo = TRUE}
intervals <- activity_valid %>% group_by(interval) %>% summarize(mean_steps = mean(steps))
with(intervals, plot(interval, mean_steps, type="l"))
paste("Time interval with maximum mean steps is",intervals$interval[which(intervals$mean_steps == max(intervals$mean_steps))])
```

## Imputing missing values

Check if any dates or intervals are missing and issue a warning if true.  
Missing step values will be imputed from the mean value for that time interval.  
Mean and median total steps per day (with missing values imputed) will be presented.

```{r impute_missing_values, echo = TRUE}
if (!anyNA(activity$date)) {
  print("No missing date values")
} else {
    print("Missing dates - use caution")
}
if (!anyNA(activity$interval)) {
  print("No missing interval values")
} else {
  print("Missing intervals - use caution")
}
activity_NA <- is.na(activity$steps)
paste("Number of missing rows is",sum(activity_NA))

activity_imputed <- activity
for (i in 1:length(activity_imputed$steps)) {
  if (activity_NA[i] == TRUE) {
    activity_imputed$steps[i] <- intervals$mean_steps[which(intervals$interval == activity_imputed$interval[i])] }
}

steps_i <- activity_imputed %>% group_by(date) %>% summarize(mean_steps = mean(steps), median_steps = median(steps))
hist(steps_i$mean_steps, main="mean steps (imputed) per day", xlab="mean steps (imputed)")
print(steps_i, n=nrow(steps_i))
```

Additional analysis: what is the average daily activity pattern when including the imputed data?

```{r daily_activity_pattern_imputed, echo = TRUE}
intervals_i <- activity_imputed %>% group_by(interval) %>% summarize(mean_steps = mean(steps))
with(intervals_i, plot(interval, mean_steps, type="l"))
paste("Time interval with maximum mean steps (imputed) is",intervals_i$interval[which(intervals_i$mean_steps == max(intervals_i$mean_steps))])
```

## Are there differences in activity patterns between weekdays and weekends?

```{r weekdays_weekends, echo = TRUE}
activity_imputed_dt <- activity_imputed
activity_imputed_dt$day_type <- NULL
for (d in 1:length(activity_imputed_dt$date)) {
  if (weekdays(as.Date(activity_imputed_dt$date[d])) %in% c("Monday","Tuesday","Wednesday","Thursday","Friday")) {
    activity_imputed_dt$day_type[d] <- "weekday"
  } else {
    activity_imputed_dt$day_type[d] <- "weekend"
    }
}

library(lattice)
intervals_dt <- activity_imputed_dt %>% group_by(day_type, interval) %>% summarize(mean_steps = mean(steps))
xyplot(mean_steps ~ interval | day_type, intervals_dt, type="l", layout=c(1,2), ylab="Mean number of steps",xlab="Interval")
```